### 2020-08-14 manuscript/manuscript.rmd
first pass to the manuscript, fix several NAs, update figure orders, fix a couple misspellings

**Fig. 2.** An overview of previous phylogenomic studieson animals. (A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister (orange nodes) or Porifera-sister (green nodes) based on their main conclusion from each study. (B) Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

@@

cat_categories$n_categories_poisson = cat_categories$`Infered categories from Poisson+CAT`
cat_categories$n_categories_gtr = cat_categories$`Infered categories from GTR+CAT`

@@

A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories. CAT models employ a Dirichlet process prior to inferring the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq]. IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from 10 (C10) to 60 (C60). Differences in analysis results could therefore be due to differences in the number of categories. The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses. This requires a very large number of additional parameters that must be estimated from the data. Despite C60 models have a better fit than site-homogeneous models in ModelFinder, the program reported that the C60 model might be overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 is probably already too high according to ModelFinder (Supplementary Table 5). These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].

@@

@@

```{r n_cat }

# Return the allocation midpoint, ie the percent of sites that are allocated to the 
# 50% of the least frequent categories
allocation_midpoint = function ( pb_summary ) {
  x = pb_summary %>% pull( count ) %>% sort(decreasing = FALSE)
  x = x/ sum(x)
  y = cumsum( x )
  return ( round( median( y ) *100, 2 ) )
}

# Parse the last sample from chain 1 of each analysis
phil_cat_c1 = 
  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_1.chain")

phil_cat60_c1 = 
  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_nCAT60_Chain_1.chain")

whel_cat_c1 = 
  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT_Chain1.chain")

whel_cat60_c1 = 
  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT60_Chain_1.chain")

# Create a single tibble with summaries of all analyses
pb_summaries = list(
  summarise_sample( phil_cat_c1 ) %>% 
    mutate ( chain=1, matrix="Philippe2009", model="Poisson+CAT"),
  summarise_sample( phil_cat60_c1 ) %>% 
    mutate ( chain=1, matrix="Philippe2009", model="Poisson+nCAT60"),
  summarise_sample( whel_cat_c1 ) %>% 
    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+CAT"),
  summarise_sample( whel_cat60_c1 ) %>% 
    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+nCAT60")
) %>%
  bind_rows() %>%
  mutate( analysis = paste(matrix, model, sep=" ")) %>%
  mutate( model=factor(model, levels=c("Poisson+nCAT60", "Poisson+CAT")) )

# Identify the midpoint of each set of allocations
mid_phil_cat = 
  pb_summaries %>% 
  filter( matrix=="Philippe2009" ) %>%
  filter( model=="Poisson+CAT" ) %>%
  allocation_midpoint()

mid_phil_ncat60 = 
  pb_summaries %>% 
  filter( matrix=="Philippe2009" ) %>%
  filter( model=="Poisson+nCAT60" ) %>%
  allocation_midpoint()

mid_whel_cat = 
  pb_summaries %>% 
  filter( matrix=="Whelan2017_strict" ) %>%
  filter( model=="Poisson+CAT" ) %>%
  allocation_midpoint()

mid_whel_ncat60 = 
  pb_summaries %>% 
  filter( matrix=="Whelan2017_strict" ) %>%
  filter( model=="Poisson+nCAT60" ) %>%
  allocation_midpoint()

n_categories_phil_cat =
  pb_summaries %>% 
  filter( matrix=="Philippe2009" ) %>%
  filter( model=="Poisson+CAT" ) %>%
  nrow()

n_categories_whel_cat =
  pb_summaries %>% 
  filter( matrix=="Whelan2017_strict" ) %>%
  filter( model=="Poisson+CAT" ) %>%
  nrow()

# Perform global MDS analysys
fit = cmdscale( dist( pb_frequencies ) ,eig=TRUE, k=2)
pb_summaries %<>% mutate( x_global = fit$points[,1], y_global = fit$points[,2] )

# Generate plots
gg_rank_phil = 
  pb_summaries %>% 
  filter( matrix=="Philippe2009" ) %>%
  ggplot() +
  geom_line(aes(x=rank, y=count, col=model)) + 
  xlab( "category rank" ) +
  ylab( "site count" ) +
  theme(legend.position="right")

gg_rank_whel = 
  pb_summaries %>% 
  filter( matrix=="Whelan2017_strict" ) %>%
  ggplot() +
  geom_line(aes(x=rank, y=count, col=model)) + 
  xlab( "category rank" ) +
  ylab( "site count" ) +
  theme(legend.position="right")

gg_mds_phil = 
 pb_summaries %>% 
  filter( matrix=="Philippe2009" ) %>% 
  ggplot() +
  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) + 
  xlim( -0.5, 0.5) +
  ylim( -0.6, 0.6) +
  xlab( "MDS1" ) +
  ylab( "MDS2" ) +
  theme(legend.position="right")

gg_mds_whel = 
 pb_summaries %>% 
  filter( matrix=="Whelan2017_strict" ) %>% 
  ggplot() +
  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +  
  xlim( -0.5, 0.5) +
  ylim( -0.6, 0.6) +
  xlab( "MDS1" ) +
  ylab( "MDS2" ) +
  theme(legend.position="right")

gg_chain = 
ggarrange(
  gg_rank_phil, 
  gg_rank_whel, 
  gg_mds_phil, 
  gg_mds_whel, 
  ncol  = 2, 
  nrow = 2, 
  labels = c("A Philippe2009", "B Whelan2017", "C", "D"), 
  common.legend = TRUE
)

There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister). It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categroies are available. This would suggest that we should favor the result under a very large number of categories, Porifera-sister. Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameter. This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister. One way to assess this is to look at the allocation of sites to equilibrium frequency categories. Each new category adds the same number of parameters, but that site can be allocated to any number of sites. If as we add categories those categories are allocated to a small fraction of sites, then the negative cost of those sites is high relative to the fraction of the data they help explain, and we should be skeptical of the results shen adding so many sites. That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009 and Whelan2017_strict matrices (Extended Data Fig. 3.5 a-b), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009 matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This is in stark contrast to nCAT60 analyses, constrained to 60 categories, that have no such long tail of rare categories (Extended Data Fig. 3.5 a-b). This suggests that support for Prorifera-sister is due to adding a very large number of parameters that explain a tiny fraction of the data.

We can further assess the distribution of equilibrium frequencies across categories when the number of categories are constrained to 50 versus when they are unconstrained and allowed to range into the hundred. It could be that the sites that are allocated to rate caetgories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain. Alternatively, when a large number of caetgories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value. We find support for this second scenario (Extended Data Fig. 3.5 d-c).

Taken together, these results indicate that when CAT models are run without constraint on the number of categories, they select a very large number of categories. A large number of these categories are allocated to a small number of sites that have small differences from other categories. Given the very large number of parameters that must be estimated for this large number of rare categories, this suggests that unconstrained CAT analyses are overparameterized. While models that accommodate site heterogeneity in equilibrium frequency are important in phylogenetic analyses, and tools such as PhyloBayes and IQtree that implement such models are critical to phylogenetic questions considered here, the present analyses suggest that there are potential problems with constrained CAT analyses that are prone to create a large number of categories for a small fraction of sites. In the context of the present question, this means we should be skeptical of Porifera-sister not only because it is obtained under such a narrow range of analysis conditions, but because the very large number of caetegories needed to recover this result are adding a large number of parameters that explain a small fraction of the data.

@@

To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 3). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE. Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 3; Supplementary Table 7). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes towards Porifera-sister, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 3). For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).

@@

For Poisson+CAT and LG in PhyloBayes, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. Since the default option in PhyloBayes does not provide the feature to calculate site-wise log likelihood for every generation, we replaced the line “int sitelogl = 0” with “int sitelogl = 1” in the file named “ReadSample.cpp” and installed PhyloBayes 4.1c so that site-wise log likelihood value can be stored to a file that ends with ".sitelogl" (via readpb –sitelogl). For each condition, we first calculated site-wise log likelihoods for each of two hypotheses (T1 and T2) using pb (via –T) and then stored site-wise log likelihood (a total number of samples for each site is 20) every ten until 300th generations, after discarding the first 100 generations using readpb (via -sitelogl -x 100 10 300). Next, we normalized site-wise log likelihood value across 20 samples for each of two hypotheses (T1 and T2) and combined normalized site-wise log likelihood values of T1 and T2 into a single file that was used to calculate gene-wise log-likelihood scores between T1 and T2 with Perl scripts from a previous study15

@@

@@

The 20 state matrices are the same across all analyses since they are not recoded. Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig. 3B) give insight into the technical variance of the inference process.

@@

The statistics presented in  Extended Data Figs. 4 and 5 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.

@@

@@

**Extended Data Fig. 3** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices). The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green). Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with differnet outgroups sampling and substitutional models. The GLS values for each gene in each data matrix are provided in Table S5. We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| $\geq$ 2) or weak ( 0 < |ΔlnL| < 2) phylogenetic signal.

```{r n_cat_render, fig.width=6, fig.height=7}

gg_chain

**Extended Data Fig. 3.5** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color). These plots are for singe sampled isolated from the end of PhyloBayes chain files. (A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundand to least abundant along the x axis. The unconstrained CAT analyses have a long tale categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data. The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses. (C,D) MDS plots for the amino acid frequencies. Each point is for a single category, the area of the point is proportional to the number of sites allocated to that category, and categories with closer points have more similar amino acid frequencies. A single MDS analysis was done for all analyses, so that the projections are the same in each plot. For both matrices, the much larger number of categories in the unconstrained CAT analyses laregely subdivide the amino acid frequency space already covered in the constrained nCAT60 analyses, and just subdivide that space into a far greater number of rare categories. This adds a considerable number of parameters for describing a similar overall space.

@@

**Extended Data Fig. 4.** (A) The two alternative hypotheses for deep animal relationships considered here. Relationships that are not part of these hypotheses are shown as unresolved polytomies. (B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses. These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03). Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point). For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding. This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.

@@

**Extended Data Fig. 5.** The subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions. These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane. The eight analyses are for two datasets (Chang and Whelan) and four coding schemes. The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6. Cells are color coded as in Supplemental Fig XXRecoidng_summary. Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister. The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here. The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy). The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.

![](figures/Figure_recoding_summary.png)
**Extended Data Fig. 6.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3. Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved). Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).

>>>>

**Fig. 2.** An overview of previous phylogenomic studies on animals. (A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister (orange nodes) or Porifera-sister (green nodes) based on their main conclusion from each study. (B) Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

@@

cat_categories$n_categories_poisson = cat_categories$`Infered categories from Poisson-CAT`
cat_categories$n_categories_gtr = cat_categories$`Infered categories from GTR-CAT`

@@

A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories. CAT models employ a Dirichlet process prior to inferring the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq]. IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from 10 (C10) to 60 (C60). Differences in analysis results could therefore be due to differences in the number of categories. The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr  %>% mean(na.rm=TRUE) %>% round(1)` categories for GTR+CAT analyses. This requires a very large number of additional parameters that must be estimated from the data. Despite C60 models have a better fit than site-homogeneous models in ModelFinder, the program reported that the C60 model might be overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 is probably already too high according to ModelFinder (Supplementary Table 5). These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].

@@

```{r n_cat, fig.width=6, fig.height=12}
# Return the allocation midpoint, ie the percent of sites that are allocated to the
# 50% of the least frequent categories
allocation_midpoint = function ( pb_summary ) {
  x = pb_summary %>% pull( count ) %>% sort(decreasing = FALSE)
  x = x/ sum(x)
  y = cumsum( x )
  return ( round( median( y ) *100, 2 ) )
}
# Parse the last sample from chain 1 of each analysis
phil_cat_c1 =
  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_1.chain")
phil_cat60_c1 =
  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_nCAT60_Chain_1.chain")
whel_cat_c1 =
  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT_Chain1.chain")
whel_cat60_c1 =
  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT60_Chain_1.chain")
# Create a single tibble with summaries of all analyses
pb_summaries = list(
  summarise_sample( phil_cat_c1 ) %>%
    mutate ( chain=1, matrix="Philippe2009", model="Poisson+CAT"),
  summarise_sample( phil_cat60_c1 ) %>%
    mutate ( chain=1, matrix="Philippe2009", model="Poisson+nCAT60"),
  summarise_sample( whel_cat_c1 ) %>%
    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+CAT"),
  summarise_sample( whel_cat60_c1 ) %>%
    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+nCAT60")
) %>%
  bind_rows() %>%
  mutate( analysis = paste(matrix, model, sep=" ")) %>%
  mutate( model=factor(model, levels=c("Poisson+nCAT60", "Poisson+CAT")) )
# Identify the midpoint of each set of allocations
mid_phil_cat =
  pb_summaries %>%
  filter( matrix=="Philippe2009" ) %>%
  filter( model=="Poisson+CAT" ) %>%
  allocation_midpoint()
mid_phil_ncat60 =
  pb_summaries %>%
  filter( matrix=="Philippe2009" ) %>%
  filter( model=="Poisson+nCAT60" ) %>%
  allocation_midpoint()
mid_whel_cat =
  pb_summaries %>%
  filter( matrix=="Whelan2017_strict" ) %>%
  filter( model=="Poisson+CAT" ) %>%
  allocation_midpoint()
mid_whel_ncat60 =
  pb_summaries %>%
  filter( matrix=="Whelan2017_strict" ) %>%
  filter( model=="Poisson+nCAT60" ) %>%
  allocation_midpoint()
n_categories_phil_cat =
  pb_summaries %>%
  filter( matrix=="Philippe2009" ) %>%
  filter( model=="Poisson+CAT" ) %>%
  nrow()
n_categories_whel_cat =
  pb_summaries %>%
  filter( matrix=="Whelan2017_strict" ) %>%
  filter( model=="Poisson+CAT" ) %>%
  nrow()
# Perform global MDS analysys
fit = cmdscale( dist( pb_frequencies ) ,eig=TRUE, k=2)
pb_summaries %<>% mutate( x_global = fit$points[,1], y_global = fit$points[,2] )
# Generate plots
gg_rank_phil =
  pb_summaries %>%
  filter( matrix=="Philippe2009" ) %>%
  ggplot() +
  geom_line(aes(x=rank, y=count, col=model)) +
  theme(legend.position="right") + ggtitle("Philippe2009_Choanimalia")
gg_rank_whel =
  pb_summaries %>%
  filter( matrix=="Whelan2017_strict" ) %>%
  ggplot() +
  geom_line(aes(x=rank, y=count, col=model)) +
  theme(legend.position="right") + ggtitle("Whelan2017_strict")
gg_mds_phil =
 pb_summaries %>%
  filter( matrix=="Philippe2009" ) %>%
  ggplot() +
  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +
  theme(legend.position="right")
gg_mds_whel =
 pb_summaries %>%
  filter( matrix=="Whelan2017_strict" ) %>%
  ggplot() +
  geom_point(aes(x=x_global, y=y_global, size=count, col=model), alpha=0.4 ) +  
  theme(legend.position="right")
gg_chain =
ggarrange(
  gg_rank_phil,
  gg_rank_whel,
  gg_mds_phil,
  gg_mds_whel,
  ncol  = 2,
  nrow = 2,
  labels = c("A", "B", "C", "D"),
  common.legend = TRUE
)

@@

There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister). It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categories are available. This would suggest that we should favor the result under a very large number of categories, Porifera-sister. Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameter. This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister. One way to assess this is to look at the allocation of sites to equilibrium frequency categories. Each new category adds the same number of parameters, but that site can be allocated to any number of sites. If as we add categories those categories are allocated to a small fraction of sites, then the negative cost of those sites is high relative to the fraction of the data they help explain, and we should be skeptical of the results shen adding so many sites. That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009 and Whelan2017_strict matrices (Extended Data Fig. 3A-B), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009 matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This is in stark contrast to nCAT60 analyses, constrained to 60 categories, that have no such long tail of rare categories (Extended Data Fig. 3A-B). This suggests that support for Porifera-sister is due to adding a very large number of parameters that explain a tiny fraction of the data.

We can further assess the distribution of equilibrium frequencies across categories when the number of categories are constrained to 50 versus when they are unconstrained and allowed to range into the hundred. It could be that the sites that are allocated to rate categories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain. Alternatively, when a large number of categories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value. We find support for this second scenario (Extended Data Fig. 3C-D).

Taken together, these results indicate that when CAT models are run without constraint on the number of categories, they select a very large number of categories. A large number of these categories are allocated to a small number of sites that have small differences from other categories. Given the very large number of parameters that must be estimated for this large number of rare categories, this suggests that unconstrained CAT analyses are overparameterized. While models that accommodate site heterogeneity in equilibrium frequency are important in phylogenetic analyses, and tools such as PhyloBayes and IQ-TREE that implement such models are critical to phylogenetic questions considered here, the present analyses suggest that there are potential problems with constrained CAT analyses that are prone to create a large number of categories for a small fraction of sites. In the context of the present question, this means we should be skeptical of Porifera-sister not only because it is obtained under such a narrow range of analysis conditions, but because the very large number of  categories needed to recover this result are adding a large number of parameters that explain a small fraction of the data.

@@

To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 4). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE. Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 4; Supplementary Table 7). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes towards Porifera-sister, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 4). For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).

@@

For Poisson+CAT and LG in PhyloBayes, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. Since the default option in PhyloBayes does not provide the feature to calculate site-wise log likelihood for every generation, we replaced the line “int sitelogl = 0” with “int sitelogl = 1” in the file named “ReadSample.cpp” and installed PhyloBayes 4.1c so that site-wise log likelihood value can be stored to a file that ends with ".sitelogl" (via readpb –sitelogl). For each condition, we first calculated site-wise log likelihoods for each of two hypotheses (T1 and T2) using pb (via –T) and then stored site-wise log likelihood (a total number of samples for each site is 20) every ten until 300th generations, after discarding the first 100 generations using readpb (via -sitelogl -x 100 10 300). Next, we normalized site-wise log likelihood value across 20 samples for each of two hypotheses (T1 and T2) and combined normalized site-wise log likelihood values of T1 and T2 into a single file that was used to calculate gene-wise log-likelihood scores between T1 and T2 with Perl scripts from a previous study [@shen2017contentious].

@@

To compare the allocation of frequency categories across sites in the Philippe2009_Choanimalia and Whelan2017_strict matrices for the constrained Posson+CAT60 model and unconstrained Posson+CAT model, we parsed the information of PhyloBayes chain files by sampling one in every 1000 generations after burnin determined above. The scripts and subsampled chain files are in the `../trees_new/frequency` subdirectory of the git repository.

@@

The 20 state matrices are the same across all analyses since they are not recoded. Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig. 5) give insight into the technical variance of the inference process.

@@

The statistics presented in  Extended Data Figs. 5 and 6 were parsed from the Feuda *et al.* manuscript into the file `tidy_table_3.tsv` and rendered for display with the code in `manuscript.rmd`.

@@

```{r n_cat_render, fig.width=6, fig.height=7}
gg_chain
```

**Extended Data Fig. 3** The allocation of frequency categories across sites in the Philippe2009 matrix (left column) and Whelan2017_strict matrix (right column) for the constrained Posson+nCAT60 model and unconstrained Posson+CAT models (differentiated with color). These plots are for singe sampled isolated from the end of PhyloBayes chain files. (A,B) The count of sites allocated to each equilibrium frequency category, with the categories ranked from the most abundand to least abundant along the x axis. The unconstrained CAT analyses have a long tale categories that are allocated to very few sites, which adds a considerable number of parameters that pertain to only a small fraction of the data. The nCAT60 analyses, which are constrained to 60 sites, have no such long tail and the rarest categories are allocated to a far greater number of sites than any of the sites in the long tail of the CAT analyses. (C,D) MDS plots for the amino acid frequencies. Each point is for a single category, the area of the point is proportional to the number of sites allocated to that category, and categories with closer points have more similar amino acid frequencies. A single MDS analysis was done for all analyses, so that the projections are the same in each plot. For both matrices, the much larger number of categories in the unconstrained CAT analyses laregely subdivide the amino acid frequency space already covered in the constrained nCAT60 analyses, and just subdivide that space into a far greater number of rare categories. This adds a considerable number of parameters for describing a similar overall space.

@@

**Extended Data Fig. 4** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice in Philippe2009, Ryan2013 and Whelan2017_full matrices (with two outgroup sampling: Choanimalia and full matrices). The two alternative topological hypotheses are: Ctenophora-sister; T1 (Orange);Porifera-sister T2 (Green). Proportions of genes supporting each of two alternative hypotheses in the Philippe2009, Ryan2013_est and Whelan2017_full data matrices with differnet outgroups sampling and substitutional models. The GLS values for each gene in each data matrix are provided in Table S5. We considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| $\geq$ 2) or weak ( 0 < |ΔlnL| < 2) phylogenetic signal.

@@

**Extended Data Fig. 5.** (A) The two alternative hypotheses for deep animal relationships considered here. Relationships that are not part of these hypotheses are shown as unresolved polytomies. (B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses. These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03). Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point). For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding. This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.

@@

**Extended Data Fig. 6.** The subset of eight GTR+CAT analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions. These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane. The eight analyses are for two datasets (Chang and Whelan) and four coding schemes. The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6. Cells are color coded as in Supplemental Fig XXRecoidng_summary. Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister. The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here. The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy). The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.

![](figures/Figure_recoding_summary.png)
**Extended Data Fig. 7.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3. Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved). Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).
